Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-31 appendix.tex
middle of revising numerical section

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} (which itself is obtained by setting
$\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy}) to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

>>>>

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used in the
adiabatic limit, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} (which itself is obtained by setting
$\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy}) to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

### 2015-03-31 linear_adia.tex
middle of revising numerical section

We remark that Eq. \ref{nearly_iso_explicit} is in the same 
form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear. The two equations
become identical after  setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and making the low-frequency, nearly-Keplerian approximation in
\citeauthor{lubow93} after a change of variables.

@@

In Appendix \ref{iso_discuss}, we collect some useful results for
isothermal  perturbations. This includes a necessary condition for
instability (there must be vertical shear) and an upper bound to the
instability growth rate (it is limited by the maximum vertical shear
rate in the disk). We also develop solutions in the thin disk
limit. The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections, and are provided for ease of reference. We refer the reader
to \cite{barker15} for a more detailed treatment of isothermal 
perturbations.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of thermal relaxation.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to dominate in numerical calculations with 
increasing $\beta$ (\S\ref{therm_relax_eff}), as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).

@@

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
For ease of analysis, we re-arrange Eq. \ref{adia_disp0} and square it
to obtain a polynomial in $\hat{\sigma}$,  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ is
complicated and generally requires a numerical solution. However,
simple solutions may be obtained in limiting cases which we consider
below.

@@

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|\epsilon q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$
is the critical thermal timescale for the fundamental mode
($M=0$). We comment that the expression for $\beta_\mathrm{crit}$ may
in fact be obtained without assuming $|\epsilon q\khat|\lesssim 1$.
We will confirm $\beta_\mathrm{crit}$ numerically in 
\S\ref{bcrit_num_test}.

@@

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}, but these
studies consider vertically localized disturbances. Our 
Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
global disturbances (which have small $M$).  We can interpret
$\beta_\mathrm{crit}$ as the thermal timescale below which
perturbations are effectively isothermal and the VSI operates.  We see
that in a thin disk, a very short thermal timescale is required
($\beta_\mathrm{crit}\ll 1$) because $\beta_\mathrm{crit}\propto \epsilon$.

>>>>

We remark that Eq. \ref{nearly_iso_explicit} is in the same 
form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear. The two equations
become identical after  setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and making the low-frequency, Keplerian approximation in
\citeauthor{lubow93} after a change of variables.

@@

In Appendix \ref{iso_discuss}, we collect some useful results for
isothermal  perturbations. This includes a necessary condition for
instability (there must be vertical shear) and an upper bound to the
instability growth rate (it is limited by the maximum vertical shear
rate in the disk). We also develop solutions in the thin disk
limit. The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections, and are provided for ease of reference. We refer the reader
to \cite{barker15} for a more detailed treatment of isothermal 
perturbations. %which also account for more general boundary
               %conditions than that considered below

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in finite thermal relaxation
timescales $\beta > 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of thermal relaxation.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the lowest order VSI mode, the `fundamental corrugation mode',
where the vertical velocity is constant throughout the disk
column. Hereafter we shall simply refer to them as the fundamental
mode.

The fundamental mode has been observed to dominate numerical simulations
\citep{nelson13,stoll14}, and are the ones we find to typically 
dominate in numerical calculations with increasing $\beta$
(\S\ref{therm_relax_eff}) for moderately large    
values of $\khat$, as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}). We will see later that consideration of low
order modes in fact provides a useful way to charactertize the effect
of increasing the thermal timescale on the VSI.

@@

Eq. \ref{nearly_iso_explicit} becomes 
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B(\hat{\sigma}) + \alpha(\hat{\sigma}) =
  M\left[A-2\alpha(\hat{\sigma})\right].\label{adia_disp0} 
\end{align}
For ease of analysis, we re-arrange Eq. \ref{adia_disp0} and square it
to obtain a polynomial in $\hat{\sigma}$,  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
The dispersion relation $\hat{\sigma}=\hat{\sigma}(\khat)$ is
complicated and generally requires a numerical solution. However,
simple results may be obtained in limiting cases which we consider
below.

@@

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. (It is important to note that marginal stability is an
assumption.) If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|\epsilon q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|\epsilon q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency inertial waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|\epsilon q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|\epsilon q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{\epsilon|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$ is the critical thermal timescale for
the fundamental mode ($M=0$). 
We comment the expression for $\beta_\mathrm{crit}$ may in fact be
obtained without assuming it is much less than unity from the onset or
$|\epsilon q\khat|\lesssim 1$. % This is because for $M=0$, 
% Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} simplify 
We will  
confirm $\beta_\mathrm{crit}$ numerically in \S\ref{bcrit_num_test}.

@@

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. 
Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
global disturbances (which have small $M$).  We can interpret 
$\beta_\mathrm{crit}$ as the thermal timescale below which 
perturbations are effectively isothermal and the VSI operates. We see
that in a thin disk, a very short thermal timescale is required
($\beta_\mathrm{crit}\ll 1$) because $\beta_\mathrm{crit}\propto
\epsilon$.

### 2015-03-31 linear_iso.tex
middle of revising numerical section

\subsection{Low-frequency and nearly-Keplerian
  approximation}\label{approx_gov} 
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. %\citep[e.g.][]{lin10}. 
  We call this the low frequency approximation. %physically inertial
                                %waves 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with
  $\Omega_k^2$. We call this the nearly-Keplerian approximation.

@@

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplanetary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion.

>>>>

\subsection{Low-frequency and Keplerian
  approximation}\label{approx_gov} 
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. %\citep[e.g.][]{lin10}. 
  We call this the low frequency approximation. We are therefore
  considering inertial waves \citep{barker15}.

\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with
  $\Omega_k^2$. We call this the Keplerian approximation.

@@

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms of $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplanetary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion.

### 2015-03-31 numerical.tex
middle of revising numerical section

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the  low-frequency, nearly-Keplerian and thin-disk 
approximations used in the analytical discussion above. We 
use the full expressions of $\rho$, $\Omega$ and 
$\kappa$ and $c_s$ given in \S\ref{eqm}. We also set $\hat{g}_c=1$ to   
account for the background radial disk structure.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness, and as a way of introducing terminology, 
we review this case before considering the effect of 
thermal relaxation. %A more detailed analysis is given by \cite{barker15}.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are inertial
waves made unstable by vertical shear and have been termed `body
modes' by \cite{barker15}.  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. Hereafter
we will simply refer to this as the fundamental mode, and it is robust
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased \citep{barker15}.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and \cite{barker15}. 
This a class of unstable modes associated with vertical shear near the
disk boundaries \citep{barker15}. They are also absent in our  
analysis because there we assumed a disk with infinite 
vertical extent, as the vertically isothermal disk has no
surface. However, surface modes can occur in the present calculations 
because there is an imposed vertical boundary.

@@

As explained by \cite{barker15}, while surface modes have a physical
origin, their occurence in the vertically isothermal disk is due to
imposing a finite vertical domain size. The fastest growing mode is a
surface mode, if they exist, but their association with vertical
boundaries means that $\mathrm{max}(\nu)$ depends on the choice of
$\zmax$. This has important implications when interpreting results
obtained from the vertically isothermal disk model. We return to this
issue later.

>>>>

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the  low-frequency, Keplerian and thin-disk 
approximations used in the analytical discussion above and 
use the full expressions of $\rho$, $\Omega$ and 
$\kappa$ and $c_s$ given in \S\ref{eqm}. We also set $\hat{g}_c=1$ to   
account for the background radial disk structure.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, already considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. 
For completeness, and as a way of introducing terminology, 
we review this case before considering the effect of 
thermal relaxation. %A more detailed analysis is given by \cite{barker15}.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citep{barker15}. (The possible exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. It is robust
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased \citep{barker15}.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack the distinct `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and \cite{barker15}. 
This a class of unstable modes associated with vertical shear near the
disk boundaries \citep{barker15}. They are absent in our  
analysis because there we assumed a disk with infinite 
vertical extent, since the vertically isothermal disk has no
surface. However, surface modes can occur in the present calculations 
because there is an imposed vertical boundary.

@@

As explained by \cite{barker15}, while surface modes have a physical
origin, their occurence in the vertically isothermal disk is due to
imposing a finite vertical domain size. This has important
implications when interpreting results obtained from the vertically
isothermal disk model. For example, seeking the most unstable mode may
result in a surface mode, but its growth rate would depend on
$\zmax$. We return to this issue later.

### 2015-03-31 paper.tex
revised intro and abstract
middle of revising numerical section

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. % with, for example, radial variations in temperature. 
  However, the VSI requires a short thermal or cooling 
  timescale in sub-adabatically (stably) stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short lengthscales $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer
  lengthscales $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/bcrit_mmsn2d} 
  \caption{Estimate of the thermal timescale  in a fiducial
    protoplanetary disk model, the Minimum Mass Solar Nebula, for
    perturbations with radial lengthscales $\sim 10^{-2}H$, where $H$
    is the disk scale-height. The red line corresponds to a 
    characteristic thermal timescale, derived in this work, 
    below which the VSI is expected to operate effectively. 
    \label{bcrit_mmsn2d} 
  }
\end{figure}

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. % with, for example, radial variations in temperature. 
  However, the VSI requires a short thermal or cooling 
  timescale in sub-adabatically (stably) stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/bcrit_mmsn2d} 
  \caption{Estimate of the thermal timescale  in a fiducial 
    protoplanetary disk model, the Minimum Mass Solar Nebula, for
    perturbations with radial lengthscales $\sim 10^{-2}H$, where $H$
    is the disk scale-height. The red line corresponds to a 
    characteristic thermal timescale, derived in this work, 
    below which the VSI is expected to operate effectively (the region
    outside the line).   
    \label{bcrit_mmsn2d} 
  }
\end{figure}
